A Certified Interpreter for the List Machine Benchmark

### 2020-06-08 papers/SBLP2020/paper.tex
Introduction finished

The development of new programming language design, linguistic construct
or type system involves its careful formalization in order to express
its core ideas in a concise way. However, such models have many details
and complexities which hinders its correctness assurances.
Because of such problems, the programming languages research community
started to use tools, like proof assistants~\cite{Stump16,Chlipala13},
and benchmark problems to validate them and stress its suitability for
such tasks~\cite{Aydemir05,Pientka18,Appel07}.

The rest of this paper is organized as follows. Section~\ref{sec:background}
presents a brief introduction to Agda and reviews the list machine benchmark.
In Section~\ref{sec:typing}, we
describe the type system for PEGs and its relation with the original
well-formedness predicate proposed by Ford~\cite{Ford2004}.
Section~\ref{sec:interpreter} describes the intrinsically typed syntax for PEGs
and its interpreter. Limitations of our approach are discussed
in Section~\ref{sec:discussion}. Related work is discussed in
Section~\ref{sec:related}, and Section~\ref{sec:conclusion} concludes.

All the source code in this article has been formalized in Agda
version 2.6.0 using Standard Library 1.0. All source code produced,
including the \LaTeX~ source of this
article, are available on-line~\cite{peg-rep}.

\section{Background}\label{sec:background}

\paragraph{An Overview of Agda}
Agda is a dependently-typed functional programming language based on
Martin-L\"of intuitionistic type theory~\cite{Lof98}.  Function types
and an infinite hierarchy of types of types, \ensuremath{\D{Set}\;\V{l}}, where \ensuremath{\V{l}} is a
natural number, are built-in. Everything else is a user-defined
type. The type \ensuremath{\D{Set}}, also known as \ensuremath{\D{Set}_{\D{0}}}, is the type of all
``small'' types, such as \ensuremath{\D{Bool}}, \ensuremath{\D{String}} and \ensuremath{\D{List}\;\D{Bool}}.  The type
\ensuremath{\D{Set}_{\D{1}}} is the type of \ensuremath{\D{Set}} and ``others like it'', such as \ensuremath{\D{Set}\;\to \;\D{Bool}}, \ensuremath{\D{String}\;\to \;\D{Set}}, and \ensuremath{\D{Set}\;\to \;\D{Set}}. We have that \ensuremath{\D{Set}\;\V{l}} is an
element of the type \ensuremath{\D{Set}\;(\V{l+1})}, for every $l \geq 0$. This
stratification of types is used to keep Agda consistent as a logical
theory~\cite{Sorensen2006}.

@@

\paragraph{The List Machine Benchmark}

@@

>>>>

The development of a new programming language design, linguistic construct
or type system involves its careful formalization in order to express
its core ideas in a concise way. However, such models have many details
and complexities which can hinders its correctness assurances.
Because of such problems, the programming languages research community
started to use tools, like proof assistants~\cite{Stump16,Chlipala13},
and benchmark problems to validate them and stress its suitability for
such specification tasks~\cite{Aydemir05,Pientka18,Appel07}.

A popular approach for specifying formal semantics is the
use of definitional interpreters, which represents the meaning of a
programming language as an interpreter written in some
meta-language~\cite{Reynolds72}. A major advantage of such approach
is the possibility of validating the semantics through execution.
Recently, definitional interpreters were used to formalize type
soundness theorems for some advanced typing features~\cite{Amin17}
and the semantics of imperative programming languages in which static
semantics is ensured by dependently-typed
syntax\footnote{Also known as intrinsically-typed.}~\cite{Poulsen18}.
In this work, we follow Poulsen et. al and use an intrinsically-typed
representation to build a definitional interpreter for a low-level virtual
machine developed by Appel and Leroy as a benchmark problem closer to
real-world implementations like typed assembly languages~\cite{CraryM99} and
proof carrying code~\cite{Necula97}.

More specifically, we contribute:

\begin{itemize}
  \item We show how all the details of the list machine type system
        can be encoded as dependently-typed syntax which avoids, by construction,
        the presence of stuck states in its definitional interpreter.
  \item We provide a provably correct implementations for testing the subtyping
        relation and to calculate the least common super type of two input
        types for the machine registers.
\end{itemize}

The rest of this paper is organized as follows. Section~\ref{sec:agda}
presents a brief introduction to Agda and Section~\ref{sec:list}
reviews the list machine benchmark and presents its syntax and type system.
We describe the intrinsically-typed representation for the list machine syntax,
the subtyping relation and the least common super type algorithm in
Section~\ref{sec:typing}. The list machine semantics and its realization as
a definitional interpreter are presented in Section~\ref{sec:semantics}.
Related work is discussed in Section~\ref{sec:related},
and Section~\ref{sec:conclusion} concludes.

All the source code in this article has been formalized in Agda
version 2.6.1 using Standard Library 1.3. All source code produced,
including the \LaTeX~source of this article, are available
on-line~\cite{list-rep}.

\section{An Overview of Agda}\label{sec:agda}

Agda is a dependently-typed functional programming language based on
Martin-L\"of intuitionistic type theory~\cite{Lof98}.  Function types
and an infinite hierarchy of types of types, \ensuremath{\D{Set}\;\V{l}}, where \ensuremath{\V{l}} is a
natural number, are built-in. Everything else is a user-defined
type. The type \ensuremath{\D{Set}}, also known as \ensuremath{\D{Set}_{\D{0}}}, is the type of all
``small'' types, such as \ensuremath{\D{Bool}}, \ensuremath{\D{String}} and \ensuremath{\D{List}\;\D{Bool}}.  The type
\ensuremath{\D{Set}_{\D{1}}} is the type of \ensuremath{\D{Set}} and ``others like it'', such as \ensuremath{\D{Set}\;\to \;\D{Bool}}, \ensuremath{\D{String}\;\to \;\D{Set}}, and \ensuremath{\D{Set}\;\to \;\D{Set}}. We have that \ensuremath{\D{Set}\;\V{l}} is an
element of the type \ensuremath{\D{Set}\;(\V{l+1})}, for every $l \geq 0$. This
stratification of types is used to keep Agda consistent as a logical
theory~\cite{Sorensen2006}.

@@

\section{The List Machine Benchmark}\label{sec:list}

@@

\section{Instrinsically-typed syntax}\label{sec:typing}

\section{A definitional interpreter}\label{sec:semantics}

\section{Related work}\label{sec:related}

\section{Conclusion}\label{sec:conclusion}
